Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-09 doc/mc-paper.tex
Progress on paper

\newcommand{\pand}{\mathbin{\land^{\sf p}}}
\newcommand{\por}{\mathbin{\lor^{\sf p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf a}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

@@

\newcommand{\assign}{\alpha}
\newcommand{\passign}{\rho}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}

@@

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\prov}{\mathit{Prov}}

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight the following special classes of Boolean formulas:
  \begin{description}
    \item[Normal form:] A formula where negation is applied only to variables.
    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
      \begin{itemize}
      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a disjunction must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
  \end{description}
     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
     A formula in conjunctive normal form can be represented by a set
     of {\em clauses}, each of which is a set of literals.  Each
     clause represents the disjunction of the literals, and the formula is the conjunction of its clauses.

@@

We propose a general framework for 
summarizing properties of Boolean formulas.

@@

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, the summation $\sum$ is computed using operation $\radd$, and the product $\prod$ is computed using operation $\rmul$.
\label{def:labeling}
\end{definition}

A wide variety of important tasks for Boolean functions can be
expressed as ring evaluation problems.  The 
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{description}
\item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
\end{description}
\end{proposition}

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
  Negation could be useful in some contexts, such as when
  converting binary decision diagrams (BDDs) having complemented
  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  disjunction node $\nodev$ in a d-DNNF can have at most two children $\nodeu_1$ and $\nodeu_0$ for which there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(u_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

  %% NNF
  \draw[thick] (20,12) [-latex] -- (28,12);
  \node [above] at (24,12) {\texttt{.nnf}};

@@

A CRAT file provides both a declaration of a POG, as well as a checkable, clausal
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG, thus enabling certified
knowledge compilation.  The proof system is based on extended
resolution~\cite{Tseitin:1983} with reverse unit propagation
(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
proving that a set of clauses logically implies another clause.

@@

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Each such declaration implicitly adds an extension variable $u$ representing each  POG node $\nodeu$ and a set of {\em defining} clauses
  encoding the conjunction or disjunction operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring
  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
     & {\tt dc} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
    \midrule
    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
    \midrule
     & {\tt r} & $L$             &            & Declare root literal\\
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.
Clauses can be added via RUP addition
(command {\tt a}), with a sequence of antecedent clauses (the
``hints'').
Similarly for clause deletion (command {\tt dc}).

The declaration of a $\pand$ operation has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
\end{tabular}
\end{center}
Symbol `{\tt p}' refers to ``product.''
Integer $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$  
integers representing literals of existing variables.  
This declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the conjunction of its arguments:
\begin{center}
\begin{tabular}{cccccc}
\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
  $i\!+\!1$ & $-v$ & $\lit_1$  \\
  $i\!+\!2$ & $-v$ & $\lit_2$  \\
  & $\ldots$ \\
  $i\!+\!k$ & $-v$ & $\lit_k$  \\
\end{tabular}
\end{center}
The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A $\pand$ operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then becomes a way
of representing constant $0$.

The declaration of a $\por$ operation has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} 
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
Symbol `{\tt s}' refers to ``sum.''
Integer $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers representing literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs.
This declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the disjunction of its arguments:
\begin{center}
\begin{tabular}{cccc}
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
  $i\!+\!1$ & $v$ & $-\lit_1$ \\
  $i\!+\!2$ & $v$ & $-\lit_2$ \\
\end{tabular}
\end{center}
The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$.

@@

The sequence of operator declarations, asserted clause, and
clause deletions represents a systematic transformation of the input formula
into a POG\@.  Validating each of these steps serves to prove that the
POG is logically equivalent to the input formula.

Observe that the defining clauses for the conjunction and disjunction
operations provide a full Tseitin encoding of the operations.  For any node $\nodeu$ and associated extension variable $u$,
we can therefore assume that any total assignment $\assign$ to the input variables will also assign
a value to the extension variables such that $\assign(u) = 1$ if and
only if $\assign \in \modelset(\nodeu)$.

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of
steps that both add and delete clauses.  Each addition must be truth
preserving, that is, any satisfying total assignment for the set of clauses
before clause addition should still be a satisfying assignment afterwards.
Each deletion must be falsehood preserving.  That is, there can be no
new satisfying assignments when the clause is deleted.

At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item All of the input clauses have been deleted.
\item There should be exactly one clause left that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\end{itemize}
The sequence of clause addition steps provide a proof that
$\modelset(\phi) \subseteq \modelset(\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$, and therefore 
$\modelset(\noder) \subseteq \modelset(\phi)$.

@@

 \texttt{dc 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{dc 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{dc 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{dc 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{dc 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\

@@

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CRAT file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing
four conjunctions and two disjunctions.  We name these by the node
type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
extension variable.  We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
  The first part of the CRAT file (C) declares
these nodes using clause IDs that increment by three or four,
depending on whether the node has two children or three.  The last two
nonzero values in the sum declarations are hints providing the
required mutual exclusion proof.

>>>>

\newcommand{\pand}{\mathbin{\land^{\sf p}}}
\newcommand{\por}{\mathbin{\lor^{\sf p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf p}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf p}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

@@

\newcommand{\assign}{\alpha}
\newcommand{\passign}{\rho}
\newcommand{\lassign}{\beta}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

\newcommand{\indegree}{\textrm{indegree}}
\newcommand{\outdegree}{\textrm{outdegree}}
\newcommand{\validate}{\textsf{validate}}
\newcommand{\prov}{\textrm{Prov}}

@@

\newcommand{\simplify}[2]{#1|_{#2}}

@@

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when 1) it is in
  normal form, and 2) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses in a formula and the
  literals in a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

  A {\em partitioned-operation formula} 
 satisfies the following for all product and sum operations: 
      \begin{itemize}
      \item The arguments to a product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a sum must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

We propose a general framework for summarizing properties of Boolean
formulas inspired by the formulation of {\em algebraic model counting}
by Kimmig, et al.~\cite{kimmig:jal:2017}.

@@

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, the summation $\sum$ is computed using addition operation $\radd$, and the product $\prod$ is computed using multiplication operation $\rmul$.
\label{def:labeling}
\end{definition}

Many important queries about Boolean formulas can be
expressed as ring evaluation problems.  The 
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{description}
\item[Negation] $\rep(\boolnot \phi, w) = \mulident - \rep(\phi, w)$
\item[Partitioned Product] $\rep\left(\Pand_{i=1,k} \phi_i, w \right) = \prod_{i=1,k} \rep(\phi_i, w)$
\item[Partitioned Sum] $\rep\left(\Por_{i=1,k} \phi_i, w\right) = \sum_{i=1,k} \rep(\phi_i, w)$
\end{description}
\end{proposition}

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  As we have seen, ring evaluation can readily handle negation, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF can have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

  %% NNF
  \draw[thick] (20,12) [-latex] -- (28,12);
  \node [above] at (24,12) {\texttt{.ddnnf}};

@@

A CRAT file provides both a declaration of a POG, as well as an easily checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG, thus enabling certified
knowledge compilation.  The proof system is based on extended
resolution~\cite{Tseitin:1983} with reverse unit propagation
(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
proving that a set of clauses logically implies another clause.

@@

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$ 
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \toprule
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause} \\
     & {\tt d} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause \\
    \midrule
    $C$    & {\tt p} & $V \; L^{*}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
    \midrule
     & {\tt r} & $L$             &            & Declare root literal\\
    \bottomrule
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.
Clauses can be added via RUP addition
(command {\tt a}), with a sequence of antecedent clauses (the
``hints'').
Similarly for clause deletion (command {\tt d}).

\begin{table}
\caption{Defining Clauses for Product (A) and Sum (B) Operations}
\begin{minipage}{0.54\textwidth} 
\begin{center}
\begin{tabular}{cccccc}
\multicolumn{6}{c}{(A).  Product Operation $\pand$}\\
\toprule
\makebox[10mm]{ID} & \multicolumn{5}{c}{Clause} \\
\midrule
  $i$ & $v$ & $-\lit_1$ & $-\lit_2$ & $\cdots$ & $-\lit_k$\\
  $i\!+\!1$ & $-v$ & $\lit_1$  \\
  $i\!+\!2$ & $-v$ & $\lit_2$  \\
  & $\ldots$ \\
  $i\!+\!k$ & $-v$ & $\lit_k$  \\
\bottomrule
\end{tabular}
\end{center}
\end{minipage}
\begin{minipage}{0.44\textwidth}
\begin{center}
\begin{tabular}{cccc}
\multicolumn{4}{c}{(B).  Sum Operation $\por$}\\
\toprule
\makebox[10mm]{ID} & \multicolumn{3}{c}{Clause} \\
\midrule
  $i$ & $-v$ & $\lit_1$ & $\lit_2$ \\
  $i\!+\!1$ & $v$ & $-\lit_1$ \\
  $i\!+\!2$ & $v$ & $-\lit_2$ \\
\bottomrule
$\;$ \\
$\;$ \\
\end{tabular}
\end{center}
\end{minipage}
\label{tab:defining}
\end{table}

The declaration of a {\em product} operation, creating a node with operation $\pand$,
 has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\tt 0} \\
\end{tabular}
\end{center}
Integer $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$  
integers representing literals of existing variables.  
As Table~\ref{tab:defining}(A) shows,
this declaration implicitly causes $k+1$ clauses to be added to the proof, defining extension variable $v$ to be the product of its arguments.

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then becomes a way
of representing constant $0$.

The declaration of a {\em sum} operation, creating a node with operation $\por$, has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{{\tt s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} 
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
Integer $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers representing literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs, all of which must be defining clauses for other POG operations.
As Table~\ref{tab:defining}(B) shows,
this declaration implicitly causes three clauses to be added to the proof, defining extension variable $v$ to be the sum of its arguments.
The hints must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.

@@

The defining clauses for a product or sum
operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
For the
extension variable $u$ associated with any POG node $\nodeu$, we can therefore
assume that any total assignment $\assign$ to the input variables will
also assign a value to $u$ such that $\assign(u) =
1$ if and only if $\assign \in \modelset(\nodeu)$.  For a POG with
root literal $r$, we can define its set of models $\modelset(P)$ as
the set of all total assignments $\assign$ such that $\assign(r) = 1$.

%% A CRAT proof follows the same general form as a QRAT dual
%% proof~\cite{bryant:cade:2021}---one that ensures that each clause
%% addition and each clause deletion preserves equivalence.  With CRAT,
%% however, clauses are defined both explicitly and implicitly.  Starting
%% with the set of input clauses, the proof consists of a sequence of
%% steps that both add and delete clauses.  Each addition must be truth
%% preserving, that is, any satisfying total assignment for the set of clauses
%% before clause addition should still be a satisfying assignment afterwards.
%% Each deletion must be falsehood preserving.  That is, there can be no
%% new satisfying assignments when the clause is deleted.

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG\@.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item There should exactly one remaining clause that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\item All of the input clauses have been deleted.
\end{itemize}
The sequence of clause addition steps provide a {\em forward implication} proof that
$\modelset(\phi) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\phi)$.

@@

 \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\

@@

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CRAT file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing
four products and two sums.  We name these by the node
type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
extension variable.
% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
  The first part of the CRAT file (C) declares
these nodes using clause IDs that increment by three or four,
depending on whether the node has two children or three.  The last two
nonzero values in the sum declarations are hints providing the
required mutual exclusion proof.

We step through a detailed example to provide a better understanding of the CRAT proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CRAT file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as hints.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses resolve to the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as hints.  These resolve to the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.
